Fail Faster

### 2025-03-17 paper/paper.tex

The performance difference the two versions of this generator (Figure~\ref{fig:overhead-explanation-perf}) is stark. In both
languages, the inlined version takes \textbf{half} as much time to generate a
single pair of \texttt{int}s.

\begin{figure}
  \begin{tabular}{lll}
  Library    & Generator          & Time per Generation (ns) \\
  \bq{}         & \texttt{int\_pair}          & 70                       \\
  \bq{}         & \texttt{int\_pair\_inlined} & 35                       \\
  ScalaCheck & \texttt{intPair}            & 458                      \\
  ScalaCheck & \texttt{intPairInlined}     & 266                     
  \end{tabular}
\label{fig:overhead-explanation-perf}
\end{figure}

% ┌───────┬──────────┬─────────┬────────────┐
% │ Name  │ Time/Run │ mWd/Run │ Percentage │
% ├───────┼──────────┼─────────┼────────────┤
% │ bq    │  70.47ns │  78.00w │    100.00% │
% │ fused │  35.86ns │  78.00w │     50.89% │
% └───────┴──────────┴─────────┴────────────┘

>>>>

The performance impact of this inling is large (Figure~\ref{fig:overhead-explanation-perf}). In both languages, the
inlined version takes (on average) \textbf{half} as much time to generate a single pair of
\texttt{int}s. 
Microbenchmarks of more realistic generators (see Section~\ref{section:eval}
\jwc{more precise ref}) show an even more dramatic performance boost for
inlining in this manner.

% Modern compilers use heuristics to decide when to specialize and inline code \cn, and these heuristics cannot guarantee that generator abstractions are zero-cost.
% These heuristics are also necessarily conservative about inlining and specializing recursive functions, which all generators of recursive data types must be.
% The

\begin{figure}
  \begin{tabular}{lll}
  Library    & Generator          & Average Time per Generation (ns) \\
  \bq{}         & \texttt{int\_pair}          & 70                       \\
  \bq{}         & \texttt{int\_pair\_inlined} & 35                       \\
  ScalaCheck & \texttt{intPair}            & 458                      \\
  ScalaCheck & \texttt{intPairInlined}     & 266                     
  \end{tabular}
\caption{Microbenchmarks of Generators in \bq{} (using \texttt{core\_bench} \cn) and ScalaCheck (using \texttt{jmh} \cn)}
\label{fig:overhead-explanation-perf}
\end{figure}
